---
title: "PM566 Midterm Project"
author: "Leona Ma"
date: "10/21/2021"
output: 
    html_document:
      toc: yes 
      toc_float: yes
      keep_md : yes 
    github_document:
      html_preview: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction (provide background on your dataset and formulated question), 
Methods (include how and where the data were acquired, how you cleaned and wrangled the data, what tools you used for data exploration), 
Preliminary Results (provide summary statistics in tabular form and publication-quality figures, take a look at the kable function from knitr to write nice tables in Rmarkdown), 
and a brief Conclusion about what you found in terms of the formulated question.

Introduction:
Methods:

```{r loading needed library}
library(data.table)
library(dplyr)
library(dtplyr)
library(plyr)
library(tidyverse)
```

# Part 1 Data Cleaning and Wrangling

## Step1 reading and combining the data sets
```{r reading and combining}
tobacco10 <- data.table::fread("tobacco2010.csv")
tobacco11 <- data.table::fread("tobacco2011.csv")
location <- data.table::fread("location.csv")
tobacco <- rbind(tobacco10, tobacco11)
tobacco11_L <- merge(tobacco11, location, by = "State")
```
tobacco11_L missed two obs, but it's fine, we don't want nationwide in map. 

## Step2 creating new varible "smokers and nonsmokers"
```{r creating}
tobacco[, smokers := `Smoke everyday`+`Smoke some days`]
tobacco[, nonsmokers := `Former smoker`+`Never smoked`]
```

## Step3 Spliting the data sets
```{r spliting}
count(tobacco, 'State')
nation <- tobacco[State == "Nationwide (States and DC)"]
nation_t <- tobacco[State == "Nationwide (States, DC, and Territories)"]
area <- tobacco[!State == "Nationwide (States, DC, and Territories)" & !State == "Nationwide (States and DC)"]
```


# Part 2 Looking at the Data (EDA)
summary of overall trend in nation wide
```{r}
summary(tobacco$smokers)
```

## Step1 Facet plot showing scatterplots with regression lines of different smoking status vs Year by State.

1. Smoke everyday
```{r}
ggplot(data=area, mapping=aes(x = Year, y = `Smoke everyday`)) +
  geom_point() + 
  facet_wrap(~State) +
  geom_smooth(method='lm', formula = y~x)+
  labs(title = "scatterplots of 'Smoke everyday' vs Year by State", x="Yeas", y="Percentage of people who smoke everyday")
```

2. Smoke some days
```{r}
ggplot(data=area, mapping=aes(x = Year, y = `Smoke some days`)) +
  geom_point() + 
  facet_wrap(~State) +
  geom_smooth(method='lm', formula = y~x)+
  labs(title = "scatterplots of 'Smoke some days' vs Year by State", x="Yeas", y="Percentage of people who smoke some days")
```

3. Former smoker
```{r}
ggplot(data=area, mapping=aes(x = Year, y = `Former smoker`)) +
  geom_point() + 
  facet_wrap(~State) +
  geom_smooth(method='lm', formula = y~x)+
  labs(title = "scatterplots of 'Smoke some days' vs Year by State", x="Yeas", y="Percentage of people who are former smoker")
```

4. Never smoked
```{r}
ggplot(data=area, mapping=aes(x = Year, y = `Never smoked`)) +
  geom_point() + 
  facet_wrap(~State) +
  geom_smooth(method='lm', formula = y~x)+
  labs(title = "scatterplots of 'Smoke some days' vs Year by State", x="Yeas", y="Percentage of people who are Never smoked")
```

5. Smokers regardless of frequency
```{r}
ggplot(data=area, mapping=aes(x = Year, y = smokers)) +
  geom_point() + 
  facet_wrap(~State) +
  geom_smooth(method='lm', formula = y~x)+
  labs(title = "scatterplots of smokers vs Year by State", x="Yeas", y="Percentage of people who are smokers")
```

6. present nonsmokers regardless of history
```{r}
ggplot(data=area, mapping=aes(x = Year, y = nonsmokers)) +
  geom_point() + 
  facet_wrap(~State) +
  geom_smooth(method='lm', formula = y~x)+
  labs(title = "scatterplots of smokers vs Year by State", x="Yeas", y="Percentage of people who are smokers")
```

```{r}
ggplot(data = nation, mapping =aes(x = Year, y=nonsmokers)) +
  geom_point() +
  geom_smooth(method='lm', formula = y~x)
```

